---
title: "flight_data_web_scrape"
output: md_document
date: "2026-02-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Scraping Capacity Numbers from Wikipedia

```{r}
library(tidyverse)
library(rvest)

# This Wikipedia page has 31 US Large Hub Airports and capacity numbers from 2015-2024
page <- read_html("https://en.wikipedia.org/wiki/List_of_the_busiest_airports_in_the_United_States")

capacity <- page %>% html_element(xpath='//*[@id="mw-content-text"]/div[2]/table[1]') %>% html_table()
colnames(capacity) <- c("Rank","Name","Airport","Cities","Metro Area","State","2024","2023","2022","2021","2020","2019","2018","2017","2016","2015")

# Select only IATA Code and The Capacity Numbers by Year
capacity <- capacity[,c(3,7:16)] %>% as.data.frame()
capacity %>% head()
```


# Bring in Flight Delay Data from 2015 to 2024

```{r}
delay <- read.csv("delay_core_2015_2024.csv")
delay <- delay %>% filter(Year%in%c("2015","2016","2017","2018","2019","2020","2021","2022","2023","2024"))
```

```{r}
top10 <- delay %>% group_by(Loc) %>%
  summarise(Total.Delays=sum(Tot.Delays)) %>%
  arrange(desc(Total.Delays))
top10$Loc[1:10]
top10
```

Top 10 Airports with the Most Total Delays from 2015-2024:

1. LGA (NY)
2. EWR (NJ)
3. SFO (CA)
4. ORD (IL)
5. JFK (NY)
6. LAX (CA)
7. LAS (NV)
8. DFW (TX)
9. BOS (MA)
10. PHL (PA)


```{r}
onlytop5 <- delay %>% filter(Loc %in% top10$Loc[c(1:5)])
ggplot(onlytop5,aes(x=Year,y=Tot.Delays,color=Loc,group=Loc)) + 
  geom_point() +
  geom_line() +
  labs(title="Total Flight Delays of Top 5 Airports with Most Delayed Operations",
       y = "Annual Delayed Operations",
       color = "Airport")
```

```{r}
#Reshaping and Joining Capacity and Delay Data
capacity_long <- capacity %>%
  dplyr::rename(Loc = Airport) %>%   
  tidyr::pivot_longer(
    cols = `2024`:`2015`,
    names_to = "Year",
    values_to = "Passengers"
  ) %>%
  dplyr::mutate(
    Year = as.integer(Year),
    Passengers = readr::parse_number(Passengers)
  )

head(capacity_long)
delay2 <- delay %>%
  dplyr::mutate(Year = as.integer(Year))

combo <- delay2 %>%
  dplyr::inner_join(capacity_long, by = c("Loc", "Year"))

dplyr::glimpse(combo)
```

```{r}
library(scales)
ggplot(combo, aes(x = Passengers)) +
  geom_histogram(bins = 30, fill = "steelblue") +
  scale_x_continuous(labels = scales::label_number(scale = 1e-6, suffix = "M")) +
  labs(title = "Distribution of Passenger Volume",
       x = "Passengers (Millions)")

ggplot(combo, aes(x = Tot.Delays)) +
  geom_histogram(bins = 30, fill = "darkorange") +
  labs(title = "Distribution of Total Delays",
       x = "Total Delays")
```
Passenger volume is right-skewed, with most airports serving between 5 and 25 million passengers and a small number of very large hub airports serving over 40 million. Total delays are highly right-skewed, indicating that while most airports experience relatively few delays, a small number of major airports account for disproportionately high delay counts. These plots provide context for the analysis by showing that both variables are dominated by a few large airports, which influences how relationships in the data should be interpreted


```{r}
#Passengers vs Total Delays
ggplot(combo, aes(x = Passengers, y = Tot.Delays)) +
  geom_point(alpha = 0.35, size = 1) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_continuous(labels = label_number(scale = 1e-6, suffix = "M")) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Passengers vs Total Delays (Airport-Year, 2015–2024)",
    x = "Passengers (Millions)",
    y = "Total Delays"
  ) +
  theme_minimal(base_size = 12)
```
Each point is one airport in one year (2015–2024). The x-axis is how many passengers that airport had that year, and the y-axis is how many total delayed flights it had. There is a positive relationship: airports with more passengers tend to have more delays, but it’s not the only factor (weather, layout, airspace, management, etc.).

```{r}
#Passengers vs Total Delay Hours
ggplot(combo, aes(x = Passengers, y = Tot.Tot.Min/60)) +
  geom_point(alpha = 0.35, size = 1) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_continuous(labels = label_number(scale = 1e-6, suffix = "M")) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Passengers vs Total Delay Hours (Airport-Year, 2015–2024)",
    x = "Passengers (Millions)",
    y = "Total Delay Hours"
  ) +
  theme_minimal(base_size = 12)
```
Same idea as the plot above, but instead of counting how many delays, this measures how much total time was lost to delays (in hours). Again, there’s a clear upward trend. Larger airports don’t just have more delays, they also lose more total time to delays. So congestion at large airports has a real cost in hours, not just in number of events.


```{r}
#Delay Causes Over Time (Top 5 Airports)
top5_codes <- delay %>%
  group_by(Loc) %>%
  summarise(TotalDelays = sum(Tot.Delays, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(TotalDelays)) %>%
  slice(1:5) %>%
  pull(Loc)

delay_causes <- delay %>%
  select(Year, Loc, Weather.Delays, Vol.Delays, Equip.Delays, Runway.Delays, Other.Delays) %>%
  pivot_longer(
    cols = ends_with(".Delays"),
    names_to = "Cause",
    values_to = "Delays"
  )
```
For the 5 most delayed airports overall (EWR, JFK, LGA, ORD, SFO), the bars show how many delays occurred each year and is broken down by cause. Weather is the dominant cause at almost all major airports and volume (congestion) is consistently the second biggest cause.

```{r}
#Congestion vs Passenger Volume
combo %>%
  mutate(vol_per_million = Vol.Delays / (Passengers / 1e6)) %>%
  ggplot(aes(x = Passengers, y = vol_per_million)) +
  geom_point(alpha = 0.35, size = 1) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_continuous(labels = label_number(scale = 1e-6, suffix = "M")) +
  scale_y_continuous(labels = comma) +
  coord_cartesian(ylim = c(0, 300)) +
  labs(
    title = "Congestion vs Passenger Volume (Zoomed)",
    x = "Passengers (Millions)",
    y = "Volume Delays per 1,000,000 passengers"
  ) +
  theme_minimal(base_size = 12)
```
This plot shows congestion-related delays per million passengers, which adjusts for airport size. The relatively flat trend suggests that although larger airports have more total delays, they are not substantially more congested on a per-passenger basis than smaller airports.






*adding
Merge Capacity and Delay Data
```{r}
#Convert Capacity dataset to long format
capacity_long <- capacity %>%
  pivot_longer(cols = -Airport, names_to = "Year", values_to = "Passengers") %>%
  mutate(
    Year = as.integer(Year),
    Passengers = as.numeric(gsub(",", "", Passengers))
  )

#Merge with Delay dataset
delay_capacity <- delay %>%
  mutate(Year = as.integer(Year)) %>%
  left_join(capacity_long, by = c("Loc" = "Airport", "Year" = "Year"))
```

Add "Delay Rate" Metric
```{r}
delay_capacity <- delay_capacity %>% 
  mutate(
    Year = as.integer(Year),
    DelayRate = Tot.Delays / Passengers * 1000000)

```
Creating DelayRate so that can interpret delays per a million passengers (since large airports will be weighted unevenly). This is to adjust for aiport size.

Delay Rate over Time
```{r}
top5_rate <- delay_capacity %>%
  group_by(Loc) %>%
  summarise(total_delays = sum(Tot.Delays, na.rm=TRUE)) %>%
  arrange(desc(total_delays)) %>%
  slice(1:5) %>%
  pull(Loc)

delay_capacity %>%
  filter(Loc %in% top5_rate, !is.na(Passengers)) %>%
  ggplot(aes(x = Year, y = DelayRate, color = Loc)) +
  geom_line(linewidth=1) +
  geom_point() +
  labs(
    title = "Delay Rate Over Time",
    y = "Delays per Million Passengers",
    x = "Year",
    color = "Airport"
  ) +
  theme_minimal()
```
Once cleaned up more, we can analyze how many delays an airport has per million passengers every year.
This is a better comparison of airports, as bigger airports will tend to have more delays.




